Density-dependent selection and the limits of relative fitness

### 2018-03-10 manuscript.tex
l

Selection is commonly described in terms of constant relative fitnesses. Yet when selection is strong, the ecological view of selection in density-regulated populations seems to be incompatible with constant-density relative fitnesses. Here we analyze the population ecological limits of relative fitness using a novel [BLAH] in which population density depends dynamically on the demographic rates of the types present. Our model contains a ``reproductive excess’’, and clearly distinguishes between density-dependent selection and selection-dependent density. These two effects are confounded in standard models of density-regulated population growth. Both effects are necessary, in combination with strong selection, for relative fitness to break down in populations close to demographic equilibrium. Remarkably, both effects are not sufficient: we give an example of strong selection on a density-regulating trait subject to density-dependent selection that conforms to the relative fitness description almost exactly. We reiterate the importance of reproductive excesses in many species, which allows even strong selection to have no effect on density. Our model also offers a possible alternative to relative fitness when the latter is untenable, as is likely the case far from demographic equilibrium.

@@

There are a variety of different measures of fitness. Some widely used examples are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, equilibrium population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where the focus is relative genotypic proportions \cite[pp. 468]{barton_2007}. The variety of fitness measures is not problematic in itself, because different measures have different uses. But it should be clear how these measures are connected to the processes of birth and death which ultimately drive selection \citep{metcalf_2007,doebeli_2017}. While such a connection is fairly clear for absolute fitness measures like $r$, relative fitness is largely divorced from population ecology. It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}. Given the ubiquitous use of relative fitness, it is important that we understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

Constant relative fitness values can be justified as an approximation which holds when selection is sufficiently weak and stable over time. Formally, these values are the ``zeroth order'' components of the actual relative fitnesses, which are not constant but depend on density, frequency, age structure, and so on \cite[pp. 277]{ewens_2004} \citep[Chap. 4]{charlesworth_1994}. Yet strong, temporally-variable selection occurs widely in nature and the lab, including in wild \textit{Drosophila}, where population density also varies by orders of magnitude each seasonal cycle \citep{messer_2016,bergland_14}. The question is whether relative fitness can be used when selection is not vanishingly weak. In general, age-structured populations that reproduce by outcrossing do not permit strong selection to be represented in terms of type-specific relative-fitness constants \citep[Chap. 4]{charlesworth_1994}. We will therefore restrict our attention to asexual haploids with little or no age structure, where it is easier to evaluate how the success or failure of the relative fitness description is tied to the underlying population ecological assumptions.

@@

The simple interpretation of relative fitness as a difference in intrinsic growth rates breaks down when we allow for population crowding. Since crowded and uncrowded conditions are so different, we expect that $s$ will often depend on density \citep{travis_2013}. Eq.~\eqref{eq:canonical} is then no longer a complete description of selection --- we would also need to specify a model for how density is changing. Note that frequency-dependent selection does not raise similar problems; Eq.~\eqref{eq:canonical} is still a complete description of selection even if its behavior is more complicated due to $s$ depending on frequency. Population genetics traditionally evades the issue of density-dependent selection by simply assuming that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant. The selection coefficient $s$ now abstractly parameterizes the rate at which selection changes relative frequencies, and no longer corresponds to differences in intrinsic growth rates $r$.

However, MacArthur famously argued that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}. The classic example is the logistic model, where the type with the greatest equilibrium population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a). Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Differences in $R^*$ will often entail differences in equilibrium density. The Lotka-Volterra competition model also couples selection in crowded populations to density except in special cases \citep{smouse_1976,mallet_2012}. In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b), even in the absence of age-structure and mating.

@@

In light of this difficulty, the relative fitness description has been justified in broadly two different ways for crowded populations (we do not discuss Wagner's [\citeyear{wagner_2010}] measure-theoretical justification, which is independent of population biology). The first is to simply assume that selection is density-independent but relax the assumption of constant $N$ by allowing density to change as a result of selective sweeps \citep[pp. 468]{barton_2007} \citep{prout_1980}. Obviously this does not address the problem that $s$ can, in reality, depend on density. Type-specific responses to density are at the center of MacArthur's argument and the density-dependent selection literature that grew out of it (e.g. \citep{roughgarden_1979}).

The second justification, which primarily grew out of a controversy over Haldane's ``cost of selection'', is to appeal to the existence of a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population,kimura1969natural,nei1971fertility}. Selection can then be concentrated at the juvenile phase, uncoupling selection from population density at the adult phase unless it is so strong that the reproductive excess is depleted. This justifies Eq.~\eqref{eq:canonical} because, for a population in demographic equilibrium, selective sweeps do not affect density, and so the density-dependence of selection does not matter. Unfortunately this reproductive excess literature is also poorly integrated with population ecology. \cite{kimura1969natural} took constant $N$ as a requirement and then derived some variants of the logistic model that satisfy this requirement. \cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types. As a result, the role of reproductive excesses in justifying Eq.~\eqref{eq:canonical} is still largely verbal.

Here we study the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests. Rather than attempting to make sense of relative fitness in existing standard models of population growth (e.g. \citep{kimura1969natural,mallet_2012}), we instead do the reverse, and attempt to make population ecological sense [BLAH]. Our starting point is the classic lottery model of territorial contest \citep{sale_77,chesson_1981}. The classic lottery assumes a saturated population with constant $N$, and fitness involves a product of fertility and juvenile viability \citep[pp. 185]{crow_1970}, [BLAH] generations can overlap. Our first task is to generalize the lottery model to create a variable-density version [BLAH] (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').

Equipped with this new model, we turn to the evaluation of Eq.~\eqref{eq:canonical}. We first discuss selection on the ability to contest territories, which behaves like a pure constant-$N$, relative fitness trait, and discuss how this fits with MacArthur's analysis of selection in crowded populations (section ``$K$-selection and selection-dependent density''). We then consider selection on density-regulating traits (section ``Density-regulating traits and the threat of strong selection''), and conclude by contrasting the classical density-dependent selection literature with our results (``Discussion'').

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} Comparison of Eq.~\eqref{eq:master}, the classic lottery model, and simulations. The vertical axis is per-propagule success rate for all propagules $\Delta_+ n_i/m_i$, and for the three separate components in Eq.~\eqref{eq:master}. Simulations are 	conducted as follows: $x_i,x_j$ values are sampled $U$ times from Poisson distributions with respective means $l_i,l_j$, and the victor in each territory is then decided by random sampling weighted by the lottery win probabilities $c_ix_i/\sum_i c_i x_i$. Two types are present, a rare type with $c_j=1.5$, and a common type with $c_i=1$. Simulation points are almost invisible for the common type due to near exact agreement with Eq.~\eqref{eq:master}. Dashed lines show the failure of the classic lottery model at low density. Parameters: $U=10^5$ and $l_j/l_i=0.1$.} 
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{DDS_lottery.pdf}
\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery plotted as the difference in propagule success rate $\Delta n_j/m_j-\Delta n_i/n_i$ between an adaptive variant $j$ and a wildtype $i$ present at the same frequency. Here $b_i=1$, $d_i=0.5$ and $c_i=1$. For $b$-selection we set $b_j=b_i(1+\epsilon)$, and similarly for $c$ and $d$, with $\epsilon=0.1$. $d$-selection is density-independent, $b$-selection gets weaker with lower territorial availability, while $c$-selection initially increases with density as territorial contests become more important, but eventually also declines as  available territories become scarce. The equilibrium density for the $i$ type is $\approx 0.4$.}
\end{figure}

@@

As shown in Fig.~\ref{fig:DDS_lottery}, $d$-selection is independent of density; the selection coefficient for a $d$-variant with $d_j= d_i(1-\epsilon)$ is a constant $s=\epsilon d_i$. Thus, to threaten the constant-$s$ assumption in an equilibrium population we require selection to act on a trait that is both density-dependent and density-regulating.

This is the case for $b$ in the variable-density lottery. However, before we discuss this relatively complicated trait, it is helpful to summarize the threat to Eq.~\eqref{eq:canonical} in simpler models of density-regulated growth, as exemplified by the birth-death model \citep{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i. \label{eq:simplebirthdeath}
\end{equation}
Here $\delta_i$ is per-capita mortality due to crowding (for simplicity, there are no deaths when uncrowded). Starting from a type $i$ population in equilibrium (we consider the non-equilibrium case in BLAH), a variant with $\delta_j=\delta_i(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_i N$ in Eq.~\eqref{eq:canonical}. This only affects the sweep if $N$ changes substantially as the population shifts to the new $j$ type equilibrium. From Eq.~\eqref{eq:simplebirthdeath} we have $N_{\rm initial}=b_i/\delta_i$ and $N_{\rm final}=b_i/(\delta_i(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$, and so $s_{\rm initial}= \epsilon b_i$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$. Consequently, substantial deviations from Eq.~\eqref{eq:canonical} requires strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}; \citealt{kimura1969natural,crow_1970}).

@@

In the variable-density lottery, equilibrium-to-equilibrium $b$-sweeps are qualitatively different from $\delta$ sweeps, because $b$ is not a simple birth rate. Greater $b$ means more propagules contesting territories, but also more territories being contested. The density-dependence of $b$-selection is represented by the term $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:bdensitydependence}. The net effect on $f(\overline{b},N)$ is precisely zero in equilibrium: in a single-type equilibrium we have $b_i/\overline{b}=1$ and so $f(\overline{b},N)=d_i$ exactly at the beginning and end of a $b$-sweep, even though $N$ increases. Strictly speaking there is some deviation in $f(\overline{b},N)$ from $d_i$ during the sweep, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (deviation from Eq.~\eqref{eq:canonical} in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in the variable-density lottery is only of order $\epsilon^2$; see Appendix D for details). Since selection must already be quite strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonical}, the density-independent model effectively applies exactly for equilibrium $b$-sweeps (however, selection is frequency-dependent because of the $1/\overline{b}$ factor).

If selection simultaneously acts on more than one trait in the variable-density lottery, then evolution in a density-regulating trait can drive changes in the strength of selection on a trait subject to density-dependent selection (Fig.~\ref{fig:multiple}). This can produce behavior reminiscent of Fig.~\ref{fig:strengthofselection}.

\begin{figure}
\centering
\includegraphics[scale=0.8]{multiple.pdf}
\caption{\label{fig:multiple} Simultaneous selection on $d$ and $c$ as predicted by Eq.~\eqref{eq:master}. Selection is not constant over the sweep.}
\end{figure}

@@

Summarizing the three traits in the variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent. This rich variety of behaviors in relation to density is quite different from that found in the classical density-dependent selection literature \citep{roughgarden_1979,christiansen_2004}. Yet despite these differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the canonical constant-$s$ relative fitness description of selection.

To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme proposed a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater equilibrium density \citep{gill_1974}. A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. Setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$, which both behave like $\delta$ in Eq.~\eqref{eq:simplebirthdeath} in that they are density-dependent and cause density to change over a sweep (although $N$ only dips transiently during an $\alpha$-sweep). Thus, strong selection is sufficient for relative fitness to break down in the classical view of density-dependent selection. Indeed, in the defense of Eq.~\eqref{eq:canonical} given by \cite{kimura1969natural}, it was assumed that $s$ will be a few percent at most. While this may be reasonable for adaptive mutations, there is no reason to expect selection on standing variation to be so weak, wild \textit{Drosophila} being an obvious counter-example \citep{bergland_14}.

@@

Reproductive excesses appear in the variable-density lottery model when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood. Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition (a ``critical age-group''; \citealt[pp. 54]{charlesworth_1994}). In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection. In this respect, the viability selection heuristics that are common in population genetics \cite[pp. 61]{gillespie_2004} actually capture an important ecological process.

We now turn to the breakdown of Eq.~\eqref{eq:canonical}. We first discuss the problem shown in Fig.~\ref{fig:strengthofselection}, which occurs when strong selection changes population density and is also density-dependent. In the variable-density lottery, this occurs if and only if types differ in more than one trait. The $c$ and $d$ traits represent the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1980}. The combination is necessary to pose a threat to Eq.~\eqref{eq:canonical}. However, the $b$ trait remarkably demonstrates that the combination is not sufficient, since the density-dependence of $b$-selection disappears over equilibrium-to-equilibrium $b$-sweeps. Thus, the simple linear models that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012,travis_2013} actually represent a complicated form of the interaction between density and selection, and their parameters confound the underlying issues.

While this is a conceptual reason to be wary of the classical density-dependent selection models, it is not clear how we should expect the trait variation in nature to align. For instance, should we expect mutations to generally affect $b$, $c$ and $d$ independently of each other, or pleiotropically such that $\delta$-like selection is prevalent? In the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule  suggests that $\delta$-like selection will be prevalent. However, for many populations consumable resources are not well-mixed. Spatial localization of consumable resources (e.g. due to restricted movement of  nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and both it and interference competition are subsumed into the competitive ability $c$, which does not affect $N$.

Relative fitness models truly break down when $N$ is far from equilibrium and selection is density-dependent (as seems likely; \citealt{travis_2013}). For example, wild \textit{Drosophila} experience large seasonal boom-bust cycles in population density coupled to strong selection that drives large swings in allele frequency \citep{bergland_14}. In this case there is no choice but to abandon relative fitness, and our model provides one potentially suitable option. Whether or not our density-dependent lottery model is a good description of \textit{Drosophila} ecology, the close connection between our model and Wright-Fisher is useful [BLAH], because drift in our model should behave broadly similarly. Thus, our model should provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations.

>>>>

[Needs revising]
Selection is commonly described in terms of constant relative fitnesses. Yet when selection is strong, the ecological view of selection in density-regulated populations seems to be incompatible with constant-density relative fitnesses. Here we analyze the population ecological limits of relative fitness using a novel of density-dependent selection which contains a ``reproductive excess’’. Our model clearly distinguishes between density-dependent selection and changes in density driven by selection. These two effects are confounded in standard models of density-regulated population growth, but both are necessary, in combination with strong selection, for relative fitness to break down in populations close to demographic equilibrium. Remarkably, both effects are not sufficient: we give an example of strong selection on a density-regulating trait subject to density-dependent selection that conforms to the density-independent relative fitness description almost exactly. We reiterate the importance of reproductive excesses in many species, which allows even strong selection to have no effect on density. Our model also offers a possible alternative to relative fitness when the latter is untenable, as is likely the case far from demographic equilibrium.

@@

There are a variety of different measures of fitness. Some widely used examples are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, equilibrium population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where the focus is relative genotypic proportions \cite[pp. 468]{barton_2007}. The variety of fitness measures is not problematic in itself, because different measures have different uses. But it should be clear how these measures are connected to the processes of birth and death which ultimately drive selection \citep{metcalf_2007,doebeli_2017}. While such a connection is fairly clear for absolute fitness measures like $r$, relative fitness is largely divorced from population ecology. It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}.

Constant relative fitness values are typically justified as an approximation that holds when selection is sufficiently weak and stable over time (\citealt{kimura1969natural}; \citealt[pp. 277]{ewens_2004}; \citealt[Chap. 4]{charlesworth_1994}) and population density is either changing either much faster or much slower than type frequencies \citep[Fig. 9.5]{otto_2011}. Formally, these constant values are the leading terms in an expansion of a potentially complicated relative fitness function which depends on density, frequency, age structure, and so on. However, knowing that strong selection can break the relative fitness approximation in toy models like the logistic \citep{kimura1969natural}, or in highly complex age-structured models \citep[Chap. 4]{charlesworth_1994} leaves important issues unresolved. For instance, must we simply abandon relative fitness in important model systems like wild \textit{Drosophila}, where strong seasonally-alternating selection occurs in unison with ``boom-bust'' cycles in population density \citep{messer_2016,bergland_14}? And if we do abandon relative fitness, are we compelled to develop customized absolute fitness models that capture the specific ecology of the system in question? Being able to assign relative fitness values to genotypes is a powerful idealization. It is therefore important that we understand the population ecological limits of this idealization when selection is not weak and demography is not stable, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

@@

This simple interpretation of relative fitness breaks down when we allow for population crowding. Since crowded and uncrowded conditions are so different, $s$ will often depend on density \citep{travis_2013}. Eq.~\eqref{eq:canonical} is then no longer a complete description of selection --- we would also need to specify a model for how density is changing. Note that frequency-dependent selection does not raise similar problems; Eq.~\eqref{eq:canonical} is still a complete description of selection even if its behavior is more complicated due to $s$ depending on frequency. This issue does not arise in many population genetics models because total population density $N$ is assumed to have reached some fixed equilibrium value. The selection coefficient $s$ then abstractly parameterizes the rate at which selection changes relative frequencies, and no longer corresponds to differences in intrinsic growth rates $r$.

Yet MacArthur famously argued that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}. The classic example is the logistic model, where the type with the greatest equilibrium population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a). Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Differences in $R^*$ will often entail differences in equilibrium density. The Lotka-Volterra competition model also couples selection in crowded populations to density except in special cases \citep{smouse_1976,mallet_2012}. In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).

@@

In light of this, the relative fitness description has been justified in broadly two different ways for crowded populations in the strong-selection, non-equilibrium demography regime. The first is to simply assume that selection is density-independent but relax the assumption of constant $N$ by allowing density to change as a result of selective sweeps \citep[pp. 468]{barton_2007} \citep{prout_1980}. Obviously this does not address the problem that $s$ can, in reality, depend on density. Type-specific responses to density are at the center of MacArthur's argument and the density-dependent selection literature that grew out of it (e.g. \citep{roughgarden_1979}). The second justification, which primarily grew out of a controversy over Haldane's ``cost of selection'', is to appeal to the existence of a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population,kimura1969natural,nei1971fertility}. Selection can then be concentrated at the juvenile phase, uncoupling selection from population density at the adult phase unless it is so strong that the reproductive excess is depleted. Selective sweeps then need not affect density, and so the density-dependence of selection does not matter. Unfortunately this reproductive excess literature is also poorly integrated with population ecology. For instance, \cite{kimura1969natural} simply took constant $N$ as a requirement and then derived some variants of the logistic model that satisfy this requirement. \cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types. As a result, the role of reproductive excesses in justifying the use of relative fitness is still largely verbal.

Here we study the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests. We restrict our attention to asexual haploids with little or no age structure, since it is then easier to evaluate how the success or failure of the relative fitness description is tied to the underlying population ecological assumptions. Our starting point is the classic lottery model which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}. The classic lottery incorporates a reproductive excess, and fitness involves a product of fertility and juvenile viability akin to the population simplest population genetic models of selection (e.g. \citealt[pp. 185]{crow_1970}). The classic lottery also assumes constant $N$, and generations can overlap. Our first task is to generalize the lottery model to create a variable-density lottery (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').

Equipped with this model, we turn to the evaluation of relative fitness. [Need to write outline, will do when structure is finalized]

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} Comparison of Eq.~\eqref{eq:master}, the classic lottery model, and simulations. The vertical axis is per-propagule success rate for all propagules $\Delta_+ n_i/m_i$, and for the three separate components in Eq.~\eqref{eq:master}. Two types are present with $c_1=1$, $c_2=1.5$ and $l_2/l_1=0.1$. Simulations are conducted as follows: $x_1,x_2$ values are sampled $U=10^5$ times from Poisson distributions with respective means $l_1,l_2$, and the victorious type in each territory is then decided by random sampling weighted by the lottery win probabilities $c_ix_i/(c_1 x_1 + c_2 x_2)$.  Simulation points are almost invisible for the common type due to near exact agreement with Eq.~\eqref{eq:master}. Dashed lines show the failure of the classic lottery model at low density.} 
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{DDS_lottery.pdf}
\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery plotted as the difference in propagule success rate $\Delta n_2/m_2-\Delta n_1/n_1$ between an adaptive variant $2$ and a wildtype $1$ with equal frequencies. Here $b_1=1$, $d_1=0.5$ and $c_1=1$. For $b$-selection we set $b_2=b_1(1+\epsilon)$, and similarly for $c$ and $d$, with $\epsilon=0.1$. $d$-selection is density-independent, $b$-selection gets weaker with lower territorial availability, while $c$-selection initially increases with density as territorial contests become more important, but eventually also declines as  available territories become scarce. The wildtype equilibrium density is $n_1/T\approx 0.4$.}
\end{figure}

@@

As shown in Fig.~\ref{fig:DDS_lottery}, $d$-selection is independent of density; the selection coefficient for a $d$-variant with $d_2= d_1(1-\epsilon)$ is a constant $s=\epsilon d_1$. Thus, to threaten the assumption of density-independent $s$ in an equilibrium population we require selection to act on a trait that is both density-dependent and density-regulating.

This is the case for $b$ in the variable-density lottery. However, before we discuss this relatively complicated trait, it is helpful to summarize the threat to Eq.~\eqref{eq:canonical} in simpler models of density-regulated growth, as exemplified by the birth-death model \citep{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i. \label{eq:simplebirthdeath}
\end{equation}
Here $\delta_i$ is per-capita mortality due to crowding (for simplicity, there are no deaths when uncrowded). Starting from a type $1$ population in equilibrium (we consider the non-equilibrium case in BLAH), a variant with $\delta_2=\delta_1(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_1 N$ in Eq.~\eqref{eq:canonical}. This only affects the sweep if $N$ changes substantially as the population shifts to the new equilibrium. From Eq.~\eqref{eq:simplebirthdeath} we have $N_{\rm initial}=b_1/\delta_1$ and $N_{\rm final}=b_1/(\delta_1(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$, and so $s_{\rm initial}= \epsilon b_1$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$. Consequently, substantial deviations from Eq.~\eqref{eq:canonical} occurs if there is sufficiently strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}; \citealt{kimura1969natural,crow_1970}).

@@

Equilibrium-to-equilibrium $b$-sweeps in the variable-density lottery are qualitatively different from $\delta$ sweeps, because greater $b$ not only means more propagules contesting territories, but also more territories being contested. Together, the net density-dependent effect on $b$-selection is zero; in Eq.~\eqref{eq:bdensitydependence}, since $b_i/\overline{b}=1$ in a single-type equilibrium, the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is exactly equal to the constant mortality rate at the beginning and end of a $b$-sweep, even though $b$ and density change. During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation constant $s$ in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in the variable-density lottery is only of order $\epsilon^2$; see Appendix D for details). Since selection must already be strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonical}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps. Note, however, that the selection coefficient for $b$-sweeps (as  defined by differences in $\Delta n_i/n_i$) does depend on frequency because of the $1/\overline{b}$ factor.

If selection acts simultaneously on more than one trait in the variable-density lottery, then evolution in a density-regulating trait can drive changes in the strength of selection on a trait subject to density-dependent selection (Fig.~\ref{fig:multiple}). This can produce behavior analogous to selection on $\delta$ in Fig.~\ref{fig:strengthofselection}.

\begin{figure}
\centering
\includegraphics[scale=0.8]{multiple.pdf}
\caption{\label{fig:multiple} Simultaneous selection on $d$ and $c$ in the variable-density lottery model as predicted by Eq.~\eqref{eq:master}. Selection is not constant over the sweep because $d$ is density-regulating and $c$ is density-dependent.}
\end{figure}

@@

Summarizing the three traits in the variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent. Yet despite these differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the density-independent relative fitness description of selection almost exactly. This behavior is quite different from that found in the classical density-dependent selection literature \citep{roughgarden_1979,christiansen_2004}.

To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme proposed a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater equilibrium density \citep{gill_1974}. A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. Setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$, which both behave like $\delta$ in Eq.~\eqref{eq:simplebirthdeath} in that they are density-dependent and cause density to change over a sweep (although $N$ only dips transiently during an $\alpha$-sweep). Thus, strong selection is sufficient for relative fitness to break down in the classical view of density-dependent selection.

@@

Reproductive excesses appear in the variable-density lottery model when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood. Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition (a ``critical age-group''; \citealt[pp. 54]{charlesworth_1994}). In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection. In this respect, the viability selection heuristics that are common in population genetics \cite[pp. 61]{gillespie_2004} actually capture an important ecological process without jumping to the full complexity of age-structured models.

We now turn to the breakdown of the density-independent relative fitness assumption which occurs when strong selection changes population density and is also density-dependent (Fig.~\ref{fig:strengthofselection}). In the variable-density lottery, this occurs if and only if types differ in more than one trait (Fig.~\ref{fig:multiple}). The $c$ and $d$ traits represent the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1980}. The combination is necessary to pose a threat to Eq.~\eqref{eq:canonical}. However, the $b$ trait remarkably demonstrates that the combination is not sufficient, since the density-dependence of $b$-selection disappears over equilibrium-to-equilibrium $b$-sweeps. Thus, the simple differential/difference equations that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012,travis_2013} actually represent a confounded form of the interaction between density and selection.

While this is a conceptual reason to be wary of the classical density-dependent selection models, it is not clear what trait variation to expect in nature. Should we expect mutations to generally affect $b$, $c$ and $d$ independently of each other, or pleiotropically such that $\delta$-like selection is prevalent? In the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule  suggests that $\delta$-like selection will be prevalent. However, for many populations consumable resources are not well-mixed. Spatial localization of consumable resources (e.g. due to restricted movement of nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and both it and interference competition are subsumed into the competitive ability $c$, which does not affect $N$.

The above findings underscore that the most serious threat to relative fitness arises from deviations from demographic equilibrium. This can dramatically alter frequency trajectories for individual sweeps (e.g. Fig. 9.5 in \cite{otto_2011}; Fig. 5 in \cite{mallet_2012}), as well as the long-term outcomes of selection in fluctuating environments \citep{lande_2009}. This would suggest that in the wild \textit{Drosophila} example mentioned in the Introduction, there is indeed no choice but to abandon relative fitness. The variable-density lottery could provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations. One convenient feature of our variable-density lottery (and lottery models in general) is a formal similarity to the Wright-Fisher model, which should facilitate the analysis of ``density-dependent drift''.
